---
title: "Hands-On Ex03"
date: "`r Sys.Date()`"
author: "Sherry Ng"
format: html
execute:
  message: false
  warning: false
editor: visual
---

## Getting started

```{r}
pacman::p_load(sf, tmap, tidyverse, readxl, readr)
```

### Importing Geospatial Data into R

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")

mpsz
```

### Import attribute data into R

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### Data preparation

```{r}
tail(popdata, n=5)
```

Check for unique() values

```{r}
unique(popdata["AG"])
```

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### Joining attribute data and geospatial data

-   mutate_at is used to create new variable or columns to the dataframe in R. This function is under the dplyr package.

-   .fun() create a list of function calls. a flexible way to generate a named list of functions.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

left join by joining popdata2020 to mpsz. If we do another way round, the map would have blank patches or it will not be a full complete sg map.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

### Export the df to rds format

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

### Choropleth Mapping Geospatial data using tmap

-   qtm(), refers to quick thematic map plot

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### creating a choropleth map by using tmap's elements

Different kinds of color scheme palette:

-   `http://www.sthda.com/english/wiki/colors-in-r#:~:text=There%20are%203%20types%20of,YlGn%2C%20YlGnBu%20YlOrBr%2C%20YlOrRd.`

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "RdPu",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

### Drawing a choropleth map using tm_fill() and tm_border()

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

### Plotting choropleth maps with built-in classification methods

`https://www.rdocumentation.org/packages/tmap/versions/3.3-3/topics/tm_fill`

1.  jenks

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

2.  equal

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

3.  kmeans

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

4.  pretty

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "pretty") +
  tm_borders(alpha = 0.5)
```

5.  quantile

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

6.  hclust

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "hclust") +
  tm_borders(alpha = 0.5)
```

7.  headtails

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "headtails") +
  tm_borders(alpha = 0.5)
```

8.  sd

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)
```

### Plotting choropleth map with customer break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### Map layouts

Map Legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

Map Style

`https://r-tmap.github.io/tmap/reference/tmap_style.html`

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("cobalt")
```

### Cartographic Furniture

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### Reset style to default

```{r}
tmap_style("white")
```

### 

Drawing small multiple choropleth maps

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

### Defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Greens",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### Creating multiple stand-alone maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mapping Spatial object meeting a selection criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

### 
