---
title: "Hands-On Ex02"
date: "`r Sys.Date()`"
author: "Sherry Ng"
format: html
execute: 
  message: false
  warning: false
editor: visual
---

## Getting Started

List of packages will be used.

```{r}
pacman::p_load(sf, tidyverse,readr,readxl,tidyr,dplyr,ggplot2)
```

<br>

### Importing Geospatial Data

-   [**MP14_SUBZONE_WEB_PL**]{.underline}, a polygon feature layer in ESRI shapefile format

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

-   [**CyclingPath**]{.underline}, a line feature layer in ESRI shapefile format

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPathGazette")
```

-   [**PreSchool**]{.underline}, a point feature layer in kml file format.

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

<br>

### Plotting the Geospatial Data

```{r}
plot(mpsz)
```

Plot only the geometry.

```{r}
plot(st_geometry(mpsz))
```

Plot using specific attribute.

```{r}
plot(mpsz["PLN_AREA_N"])
```

<br>

### Working with Projection

Change EPSG code from 9001

```{r}
st_crs(mpsz)
```

to 3414

```{r}
mpsz3414 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

<br>

Transforming the projection of preschool from wgs84 to svy21.

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

Check:

```{r}
st_geometry(preschool3414)
```

<br>

### Importing aspatial data

read in csv file (rmb not to push the csv raw data file to github)

```{r}
listings <- read.csv("data/aspatial/listings.csv", sep=',')
```

%\>% is call the "pipe" operator. It is used to perform multiple functions.

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

<br>

### Buffering

[**nQuadSegs**]{.underline} refers to how many segments are generated per quadrant and features. Distance is in meters.

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

calculating the area

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

sum() the total are

```{r}
sum(buffer_cycling$AREA)
```

<br>

### Point in polygon count

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

\**Note that the count is slightly different from the website. I believed that it is due to the difference in number of rows in the dataset. Currently there is 1925 rows.*

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

### Density of pre-school by planning subzone

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

```{r}
plot(mpsz3414['PreSch Density'])
```

### Exploratory Data Analysis

```{r}
hist(mpsz3414$`PreSch Density`)
```

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
y <- as.numeric(mpsz3414$`PreSch Count`)
x <- as.numeric(mpsz3414$`PreSch Density`)
plot(x, y, main = "RS between Pre-school Density and Pre-school Count",
     xlab = "Pre-school density (per km sq)", ylab = "Pre-school count",
     pch = 19, frame = FALSE)
abline(lm(y ~ x, data = mpsz3414), col = "blue")
```
