---
title: "In-Class Ex02"
date: "`r Sys.Date()`"
author: "Sherry Ng"
format: html
execute: 
  message: false
  warning: false
editor: visual
---

## Getting started

Download 3 R packages:

```{r}
pacman::p_load(sf, tidyverse, funModeling)
```

<br>

Notes to take note of:

-   st\_ refers to spatial and temporal

-   sf, the table with feature attributes and feature geometries, which contains

-   sfc , the list-column with the geometries for each feature (record), which is composed of

-   sfg, the feature geometry of an individual simple feature.

## Importing Geospatial

### The geoBoundaries dataset

```{r}
geoNGA <- st_read("data/geospatial/",
                  layer = "geoBoundaries-NGA-ADM2" ) %>%
  st_transform(crs=26392)
```

<br>

### The NGA dataset

```{r}
NGAadm <- st_read("data/geospatial/",
                  layer = "nga_admbnda_adm2_osgof_20190417" ) %>%
  st_transform(crs=26392)
```

### Importing Aspatial data

```{r}
wp_nga <- read_csv("data/aspatial/WPdx.csv") %>%
  filter(`#clean_country_name` == "Nigeria")
```

### Write the extracted data into rds format

```{r}
#saveRDS(wp_nga, file = "data/rsd/wp_nga.RDS") 
```

## Converting Aspatial Data into Geospatial

-   st_as_sfc, convert foreign geomerty object to a sfc object.

-   st_as_sf, convert foreign object to an sf object. if dataset contains a column call "geometry", coords will results in overwriting of this column by sfc geometry list column.

```{r}
wp_nga$Geometry = st_as_sfc(wp_nga$`New Georeferenced Column`)
wp_nga
```

-   st_sf, is to create sf object, which extends df-like objects with a simple feature list column.

```{r}
wp_sf <- st_sf(wp_nga, crs=4326)
wp_sf
```

### Project transformation

check projection

```{r}
st_crs(wp_sf)
```

```{r}
wp_sf <- wp_sf %>%
  st_transform(crs = 26392)
```

```{r}
st_crs(wp_sf)
```

### Excluded redundant fields

```{r}
NGA <- NGAadm %>%
  select(c(3:4, 8:9))
```

### Check for duplicate names

```{r}
NGA$ADM2_EN[duplicated(NGA$ADM2_EN)==TRUE]
```

```{r}
NGA$ADM2_EN[94] <- "Bassa, Kogi"
NGA$ADM2_EN[95] <- "Bassa, Plateau"
NGA$ADM2_EN[304] <- "Ifelodun, Kwara"
NGA$ADM2_EN[305] <- "Ifelodun, Osun"
NGA$ADM2_EN[355] <- "Irepodun, Kwara"
NGA$ADM2_EN[356] <- "Irepodun, Osun"
NGA$ADM2_EN[519] <- "Nasarawa, Kano"
NGA$ADM2_EN[520] <- "Nasarawa, Nasarawa"
NGA$ADM2_EN[546] <- "Obi, Benue"
NGA$ADM2_EN[547] <- "Obi, Nasarawa"
NGA$ADM2_EN[693] <- "Surulere, Lagos"
NGA$ADM2_EN[694] <- "Surulere, Oyo"

```

```{r}
freq(data=wp_sf,
     input = "#status_clean")
```

```{r}
wp_sf_nga <- wp_sf %>%
  rename (status_clean = '#status_clean') %>%
  select (status_clean) %>%
  mutate(status_clean = replace_na(
    status_clean, "unknown"))
```

### Extracting Water Point Data

```{r}
wp_functional <- wp_sf_nga %>% 
  filter(status_clean %in%
    c("Functional",
    "Functional but not in use",
    "Functional but needs repair"))
```

```{r}
wp_nonfunctional <- wp_sf_nga %>% 
  filter(status_clean %in%
    c("Abandoned/Decommissioned",
      "Abandoned",
      "Non-Functional due to dry season",
      "Non-Functional",
      "Non functional due to dry season"))
```

extract water point with unknown status.

```{r}
wp_unknown <- wp_sf_nga %>% 
  filter(status_clean=="unknown")
```

```{r}
NGA_wp <- NGA %>%
  mutate(`total_wp` = lengths(
    st_intersects (NGA, wp_sf_nga))) %>%
  mutate (`wp_functional` = lengths(
    st_intersects(NGA, wp_functional))) %>%
  mutate (`wp_nonfunctional` = lengths(
    st_intersects(NGA, wp_nonfunctional))) %>%
  mutate(`wp_unknown` = lengths(
    st_intersects(NGA, wp_unknown)))
```

### Saving the analytical data in rds format

```{r}
#saveRDS(NGA_wp,"data/rsd/NGA_wp.rds")
```

### Visualising attributes by using statistical graphs

```{r}
ggplot(data = NGA_wp,
       aes(x = total_wp)) + 
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue") +
  geom_vline(aes(xintercept=mean(
    total_wp, na.rm=T)),
    color="red",
    linetype="dashed", size=0.8) +
  ggtitle("Distribution of total water points by LGA") +
  xlab("No. of water points") + 
  ylab("No. of\nLGAs") +
  theme(axis.title.y=element_text(angle = 0))
```
