---
title: "In-Class Ex05"
date: "06 Febuary 2023"
date-modified: "`r Sys.Date()`"
format: html
execute: 
  echo: true
  eval: true
  message: false
  warning: false
editor: visual
---

## Getting started

-   sfdep - working with simple features. point pattern analysis

-   only focus on local clq

```{r}
pacman::p_load(tidyverse, sf, tmap, sfdep)
```

### Read in the data

```{r}
studyArea <- st_read(dsn = "data",
                     layer = "study_area") %>% 
  st_transform(crs= 3829)
```

```{r}
store <- st_read(dsn = "data",
                     layer = "stores") %>% 
  st_transform(crs= 3829)
```

## Visualizing the sf layer

tm_polygon must be on top followed by the dots

```{r}
tmap_mode("view") 
tm_shape(studyArea) + 
  tm_polygons() +
tm_shape(store)+ 
  tm_dots(col = "Name", 
          size = 0.01,
          border.col = "black", 
          border.lwd = 0.5) +
  tm_view(set.zoom.limits = c(12,16))
```

## Local Colocation Quotient

number 6 = 6 neighbors (always go for even number)

```{r}
nb <- include_self(
  st_knn(st_geometry(store), 6)
)

```

```{r}
wt <- st_kernel_weights(nb, 
                        store, 
                        "gaussian", 
                        adaptive = TRUE)
```

```{r}
FamilyMart <- store %>% 
  filter(Name == "Family Mart")
A <- FamilyMart$Name
```

```{r}
SevenEleven <- store %>% 
  filter(Name == "7-Eleven")
B <- SevenEleven$Name
```

```{r}
LCLQ <-local_colocation(A,B,nb,wt,49)
```

the code below is to combine both dataframe together. in this case we cannot use relational join because they dont have unique id. LCLQ join to store

```{r}
LCLQ_stores<-cbind(store, LCLQ)
```

```{r}
tmap_mode("view") 
tm_shape(studyArea) +   
  tm_polygons() + 
tm_shape(LCLQ_stores) +   
  tm_dots(col = "X7.Eleven",           
          size = 0.1,           
          border.col = "black",           
          border.lwd = 0.5) + 
  tm_dots(col = "X7.Eleven",           
        size = 0.1,           
        border.col = "black",           
        border.lwd = 0.5) +
tm_view(set.zoom.limits = c(12,16))
```
